\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-08-17 paper.tex
Missing typo
review for removing time-independant features out of time-to-event model

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$. We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independent features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independent regularizations are used during inference, see~\eqref{eq:pen-log-lik}. Corresponding each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]

@@

In many practical applications, subjects show highly nonlinear longitudinal trajectories. In~\eqref{eq:log-lik}, the complete longitudinal history is required for the computation of the survival function. Hence, in order to produce a good estimate of $\cM_k(t)$, we consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely 
\[u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top\]
with $\alpha \in \N_+$.
% TODO: en pratique essayer d'ajouter X dans u^l(t) pr voir si ça améliore les perf ou non. Si oui l'ajouter au modèle.
The idea here is to allow a wide range of polynomial orders for the representation so that a suitable one can be automatically chosen for each trajectory -- depending on its inherent complexity -- thanks to the regularization strategy proposed in~\eqref{eq:pen-log-lik}.
We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect. Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for all $l=1, \ldots, L$.

@@

In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
\begin{equation}
  \label{eq:pen-log-lik}
	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
\end{equation}
where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{3,k})^\top \in \R_+^3$. Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
\[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
for any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l} \] for the sparse group lasso penalty with $\norm{z}_{2,l}=\norm{z^l}_2$, for which we do not have to account for group sizes since they all have the same one. 
Hence, the resulting optimization problem is written
\begin{equation}
  \label{eq:optim-pb}
   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
 \end{equation}

@@

\paragraph*{Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde-1-2}
  \left\{
    \begin{array}{ll}
        \tilde g_{i, s, \gamma_k}^1&: b_i \mapsto \exp \big\{\tilde g_{s, \gamma_k}^2(b_i) \big\}, \\
        \tilde g_{s, \gamma_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l).
    \end{array}
  \right. 
\end{equation}
We then present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:QNMCEM}) with every increase of $N$.

Concerning the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
\begin{equation*}
  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]
\end{equation*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$. The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-beta_k}
  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k).
\end{equation}

The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{align}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{align}
for all $l = 1, \ldots, L$. The closed-form update is then given by

\begin{equation}
  \label{eq:update_beta}
  {\beta_k^l}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} (y_i^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} U_{il} \Big]}.
\end{equation}

Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k}^1$ defined in~\eqref{eq:g-tilde-1-2}. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{3,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q-2}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

>>>>

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
the history of the true unobserved longitudinal process up to time $t$. For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}. \]

@@

In many practical applications, subjects show highly nonlinear longitudinal trajectories. In~\eqref{eq:log-lik}, the complete longitudinal history is required for the computation of the survival function. Hence, in order to produce a good estimate of $\cM_k(t)$, we consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely 
\[u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top\]
with $\alpha \in \N_+$.
The idea here is to let the practitioner choose a suitable polynomial order $\alpha$ for the representation -- that could also be tuned automatically using a model selection procedure and AIC~\citep{akaike1974new} for instance.
We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect. Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for all $l=1, \ldots, L$.

@@

In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
\begin{equation}
  \label{eq:pen-log-lik}
	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
\end{equation}
where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})^\top \in \R_+^2$. Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
\[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
for any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l} \] for the sparse group lasso penalty with $\norm{z}_{2,l}=\norm{z^l}_2$, for which we do not have to account for group sizes since they all have the same one. 
Hence, the resulting optimization problem is written
\begin{equation}
  \label{eq:optim-pb}
   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
 \end{equation}

@@

\paragraph*{Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
Let us introduce the following useful notations
\begin{equation}
  \label{eq:g-tilde-1-2}
  \left\{
    \begin{array}{ll}
        \tilde g_{i, s, \gamma_k}^1&: b_i \mapsto \exp \big\{\tilde g_{s, \gamma_k}^2(b_i) \big\}, \\
        \tilde g_{s, \gamma_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l).
    \end{array}
  \right. 
\end{equation}
We then present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:QNMCEM}) with every increase of $N$.

Concerning the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
\begin{equation*}
  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]
\end{equation*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$. The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-beta_k}
  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k).
\end{equation}
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{align}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{align}
for all $l = 1, \ldots, L$. The closed-form update is then given by

\begin{equation}
  \label{eq:update_beta}
  {\beta_k^l}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} (y_i^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} U_{il} \Big]}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k}^1$ defined in~\eqref{eq:g-tilde-1-2}. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q-2}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.
